---
title: "1.4 Transformations between coordinate systems"
author: "Manual of Applied Spatial Ecology"
date: "1/8/2018"
output: 
  pdf_document: default
  html_document: default
---
1\. Exercise 1.4 - Download and extract zip folder into your preferred location

2\. Set working directory to the extracted folder in R under Session - Set Working Directory...

3\. First we need to load the packages needed for the exercise
```{r include=FALSE}
library(rgdal)
```
4\. Now open the script "MDprojections.R" and run code directly from the script
```{r}
study.states<-readOGR(dsn=".",layer="MDcounties")
plot(study.states, col="grey")
```
Let's zoom into the region we have locations instead of county level
```{r}
study.zoom<-readOGR(dsn=".",layer="MDzoom")
plot(study.zoom, col="grey")
```
5\. Import the csv file that contains all the mule deer locations by ID
```{r}
muleys <-read.csv("muleysexample.csv", header=T)
```
6\. Create a spatial data frame of raw mule deer locations with projection defined similar to study site shapefile (i.e., WGS84)
```{r}
coords<-data.frame(x = muleys$Long, y = muleys$Lat)
crs<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
plot(coords)
```
7\. Remove outlier locations
```{r}
newmuleys <-subset(muleys, muleys$Long > -110.50 & muleys$Lat > 37.3 
   & muleys$Long < -107)
muleys <- newmuleys
```
8\. Create a new spatial data frame of mule deer locations with outliers removed and projection defined similar to study site shapefile (i.e., WGS84)
```{r}
coords<-data.frame(x = muleys$Long, y = muleys$Lat)
crs<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0"
plot(coords)
```

9\. Create a spatial points data frame of mule deer locations with projection defined similar to study site shapefile (i.e., WGS84)
```{r}
deer.spdf <- SpatialPointsDataFrame(coords= coords, data = muleys, proj4string = CRS(crs))
deer.spdf[1:5,]
class(deer.spdf)
proj4string(deer.spdf)
plot(deer.spdf)
points(deer.spdf, col="yellow")
```

10\. Now let's project both the mule deer locations and study site shapefile to NAD83 UTM Zone 12 (Fig. 1.4, 1.5)
```{r}
new.crs <-CRS("+proj=utm +zone=12 +datum=WGS84")
MDzoomUTM12 <-spTransform(study.zoom, CRS=new.crs)
plot(MDzoomUTM12, col="bisque")
class(MDzoomUTM12)
proj4string(MDzoomUTM12)
summary(MDzoomUTM12)

#projection for mule deer locations
deer.crs <-CRS("+proj=utm +zone=12 +datum=WGS84")
deerUTM12 <-spTransform(deer.spdf, CRS=deer.crs)
points(deerUTM12, col="red")
class(deerUTM12)
proj4string(deerUTM12)
deerUTM12[1:5,]

#See new projected coordinates in UTM 12N for the first 5 locations
coordinates(deerUTM12)[1:5,]

#plot coordinates in Lat Long over coordinates in UTM 12N
plot(coords)
par(new=TRUE)
plot(deerUTM12, col="red")

plot(study.zoom)
par(new=TRUE)
plot(deer.spdf, col="red")

plot(MDzoomUTM12,col="bisque")
par(new=TRUE)
plot(deerUTM12, col="red")
```

