---
title: "3.3 Distance Between Locations"
author: "Manual of Applied Spatial Ecology"
date: "1/21/2019"
output: 
  pdf_document: default
html_document: default
---
Determining the distance between locations or between locations and respective habitat types can serve a variety of purposes. Several resource selection procedures require a description of the daily movement distance of an animal to determine the habitat available to an animal or when generating random locations around known locations. We will start here with a method to determine the average distance moved by mule deer in Colorado in a study to determine methods to alleviate depradation on sunflowers that have become a high commodity crop in the area.

1\. Exercise 3.3 - Download and extract zip folder into your preferred location

2\. Set working directory to the extracted folder in R under Session - Set Working Directory...

3\. Now open the script "DistanceUniqueBurst.Rmd" and run code directly from the script

4\. First we need to load the packages needed for the exercise
```{r warning=FALSE,message=FALSE}
library(adehabitatLT)
library(chron)
library(class)
```
5\. Code to read in dataset then subset for an individual animal
```{r}
muleys <-read.csv("DCmuleysedited.csv", header=T)
#Code to select an individual animal
muley15 <- subset(muleys, id=="D15")
table(muley15$id)
muley15$id <- droplevels(muley15$id)

#Sort data to address error in code and then look at first 20 records of data to confirm
muley15 <- muley15[order(muley15$GPSFixTime),]
#Run code to display the first 20 records to look at what sorting did to data

```
6\. Prepare data to create trajectories using the ltraj command in Adehabitat LT
```{r}
######################################################
## Example of a trajectory of type II (time recorded) with conversion of the date to the format POSIX that nNeeds to be done to get proper digits of date into R then POSIXct uses library(chron)
da <- as.character(muley15$GPSFixTime)
da <- as.POSIXct(strptime(muley15$GPSFixTime,format="%Y.%m.%d %H:%M:%S"))
head(da)
#Attach da to muley15
muley15$da <- da

timediff <- diff(muley15$da)
muley15 <-muley15[-1,]
muley15$timediff <-as.numeric(abs(timediff)) 

#Clean up muley15 for outliers
newmuleys <-subset(muley15, muley15$X > 599000 & muley15$X < 705000 & muley15$Y > 4167000 & muley15$timediff < 14401)
muley15 <- newmuleys
```
7\. Create a spatial data frame of locations for muley 15 for use in creating trajectories that includes time difference between locations and dates in proper format (as.POSIXct)
```{r}
data.xy = muley15[c("X","Y")]
#Creates class Spatial Points for all locations
xysp <- SpatialPoints(data.xy)
proj4string(xysp) <- CRS("+proj=utm +zone=12 +ellps=WGS84")

#Creates a Spatial Data Frame from 
sppt<-data.frame(xysp)
#Creates a spatial data frame of ID
idsp<-data.frame(muley15[2])
#Creates a spatial data frame of dt
dtsp<-data.frame(muley15[24])
#Creates a spatial data frame of Burst
busp<-data.frame(muley15[23])
#Merges ID and Date into the same spatial data frame
merge<-data.frame(idsp,dtsp,busp)
#Adds ID and Date data frame with locations data frame
coordinates(merge)<-sppt
plot(merge)
```
8\. Create an object of class "ltraj" for muley15 dataset
```{r}
ltraj <- as.ltraj(coordinates(merge),merge$da,id=merge$id)
plot(ltraj)
ltraj

#Now let's look at time differences between locations before moving forward
summary(muley15$timediff)

```
9\. Need to create separate "bursts" for each trajectory based on the number of locations collected each day. In our case it was 8 (i.e., locations collected every 3 hours during a 24-hour period).
```{r}
#We want to study the trajectory of the day at the scale of the day. We define one trajectory 
#per day. The trajectory should begin at 2200 hours so the following function returns TRUE if
#the date is time between 06H00 and 23H00 (i.e. results in 7-8 locations/day bursts)
foo <- function(date) {
da <- as.POSIXlt(date)
ho <- da$hour + da$min
return(ho>15.9&ho<23.9)
}
deer <- cutltraj(ltraj, "foo(date)", nextr = TRUE)

#Notice that the above code will remove 345 relocations that fall
#outside of your time criteria
#Warning message:
#In cutltraj(ltraj, "foo(date)", nextr = TRUE) :
#  At least 3 relocations are needed for a burst
# 345 relocations have been deleted
head(deer)
```
10\. Code to change ltraj to a data.frame to summarize distance between locations for each daily burst
```{r eval=FALSE}
dfdeer <- ld(deer)
head(dfdeer)

#Code to get mean distance moved for each burst
dfdeer <- subset(dfdeer, !is.na(dfdeer$dist))#remove NAs from last location of a burst
mean_dist <- do.call(data.frame, aggregate(dfdeer$dist, by=list(dfdeer$burst), 
    function(x) c(mean = mean(x), sd = sd(x), n=abs(length(x)))))

#Write.table gives csv output of Summary.  Be sure to specify the directory 
#and the output files will be stored there 
write.table(mean, file = "Distance.csv", sep =",", row.names = TRUE, 
  col.names = TRUE, qmethod ="double")
```
