---
title: "3.1 Importing datasets from a web source"
author: "Manual of Applied Spatial Ecology"
date: "3/11/2022"
output: 
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---
Movebank.org is a cloud-based repository for relocation data from GPS-collared or VHF-collared animals. It provides a storage facility in the cloud that can serve as a backup for your data or a transfer portal to share data among colleagues or interested researchers. Similar to any email account, each user has a Movebank account that has a login and password to gain access to your data. Administration privileges can be given to anyone with an account for viewing and downloading data.

1\. Exercise 3.1 - Download and extract zip folder into your preferred location

2\. Set working directory to the extracted folder in R under Session - Set Working Directory...

3\. Now open the script "DMAdeerMovebank.Rmd" and run code directly from the script

4\. First we need to load the packages needed for the exercise
```{r warning=FALSE, message=FALSE}
library(move)
library(RCurl)
library(circular)
```
5\. Next we are going to the [Movebank](https://www.movebank.org/) home page and explore what it has to offer. Need to create an account or select a dataset that does not require permission to use.
```{r eval=FALSE}
login <- movebankLogin(username="wdwalter", password="xxxxxxx")
```
```{r eval=FALSE}
deer <- getMovebankData(study="DMA White-tailed Deer 2018 Pennsylvania USA",
                        login=login, moveObject=TRUE)
n.indiv(deer)
n.locs(deer)
#Plot the first deer in the stack
plot(deer[[1]])

#Now we will select a single deer to explore more
deer1 <- deer[['X20212_20242F']]
plot(deer1)

#Select and plot locations of the initial 2 deer in your list
deer2 <- deer[[c(1,2)]]
plot(deer2)
#Determine names of initial 2 deer selected above
namesIndiv(deer2)
```