---
title: "1.8 Creating a Hexagonal Polygon Grid Over a Study Area"
author: "Manual of Applied Spatial Ecology"
date: "3/11/2022"
output: 
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---
1\. Exercise 1.8 - Download and extract zip folder into your preferred location

2\. Set working directory to the extracted folder in R under Session - Set Working Directory...

3\. Now open the script "GridScripts.Rmd" and run code directly from the script

4\. First we need to load the packages needed for the exercise
```{r warning=FALSE, message=FALSE}
library(rgdal)
library(rgeos)
library(raster)
```
5\. Now let's have a separate section of code to include projection information we will use throughout the exercise. In previous versions, these lines of code were within each block of code
```{r warning=FALSE, message=FALSE}
crs<-"+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
Albers.crs <-CRS("+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs")
```
6\. Also need to import several shapefiles for mule deer from Section 1.3
```{r warning=FALSE, message=FALSE, hide=TRUE}
study.counties<-readOGR(dsn=".",layer="MDcounties", verbose = FALSE)
class(study.counties)#Shows class and package used
proj4string(study.counties)#Shows projection information
plot(study.counties)#plots study sites on map
study.counties@data$StateCO#Displays labels for counties

#Labels each county with @plotOrder of each polygon (i.e., county)
text(coordinates(study.counties), labels=sapply(slot(study.counties, "polygons"),
  function(i) slot(i, "ID")), cex=0.8)

muleys <-read.csv("muleysexample.csv", header=T)

#Remove outlier locations
newmuleys <-subset(muleys, muleys$Long > -110.50 & muleys$Lat > 37.8 & muleys$Long < -107)
muleys <- newmuleys
```
7\. Identify the columns with coordinates then make a spatial data frame of locations after removing outliers
```{r fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
coords<-data.frame(x = muleys$Long, y = muleys$Lat)
head(coords)
plot(coords)

deer.spdf <- SpatialPointsDataFrame(coords= coords, data = muleys, proj4string = CRS(crs))
class(deer.spdf)
proj4string(deer.spdf)
points(deer.spdf,col="red")
```
\newline
8\. Rename labels by county name otherwise plot order would be used because duplicate counties within each state (i.e., CO, UT) occured in original shapefile from ArcMap
```{r fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
row.names(study.counties)<-as.character(study.counties$StateCO)
names.polygons<-sapply(study.counties@polygons, function(x) slot(x,"ID")) 
#Now add labels of State and County to Map
plot(study.counties)
text(coordinates(study.counties), labels=sapply(slot(study.counties, "polygons"),
  function(i) slot(i, "ID")), cex=0.3)
```
\newline
9\. Now lets extract counties within the extent of the mule deer locations
```{r fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
int <- gIntersection(study.counties,deer.spdf)#requires rgeos library
clipped <- study.counties[int,]
MDclip <- as(clipped, "SpatialPolygons")
plot(MDclip,pch=16)
#Now add labels of State and County to Map
text(coordinates(MDclip), labels=sapply(slot(MDclip, "polygons"), 
  function(i) slot(i, "ID")), cex=0.8)
bbox(MDclip)
```
10\. We also can create a hexagonal grid across the study site
```{r fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
HexPts <-spsample(MDclip,type="hexagonal", n=1000, offset=c(0,0))
HexPols <- HexPoints2SpatialPolygons(HexPts)
#proj4string(HexPols) <- CRS(crs)
plot(HexPols)
```
\newline
11\. Create this hexagonal grid across our study site by zooming into deer locations from Section 1.3
```{r hide=TRUE, fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
#Import the study site zoomed in shapefile
study.zoom<-readOGR(dsn=".",layer="MDzoom")
plot(study.zoom,pch=16)
points(deer.spdf,col="red")

#Create new hexagonal grid
HexPts2 <-spsample(study.zoom,type="hexagonal", n=500, offset=c(0,0))
HexPols2 <- HexPoints2SpatialPolygons(HexPts2)
proj4string(HexPols2) <- CRS(crs)
plot(HexPols2, add=T)
#Now add labels to each hexagon for unique ID
text(coordinates(HexPols2), labels=sapply(slot(HexPols2, "polygons"), 
  function(i) slot(i, "ID")), cex=0.3)
```
\newline
12\. We can intersect the mule deer locations with the polygon shapefile (i.e., county) they occured in if needed
```{r fig.height=4, fig.width=4}
o = over(deer.spdf,study.counties)
new = cbind(deer.spdf@data, o)

#Used to rename labels by hexagonal grid ID only otherwise plot order with "IDxx" would be used
#and would throw an error (i.e., ID2, ID3)
row.names(HexPols2)<-as.character(HexPols2@plotOrder)
names.hex<-sapply(HexPols2@polygons, function(x) slot(x,"ID"))
```
13\. As an aside, we can explore how to assign the area a location occurs in by intersecting points within the polygon shapefile.
```{r fig.height=4, fig.width=4}
o2 = over(deer.spdf,HexPols2)
new2 = cbind(deer.spdf@data,o2)
HexPols2

#Now plot with new grid IDs
plot(study.zoom,pch=16)
points(deer.spdf,col="red")
plot(HexPols2, add=T)
#Now add labels of State and County to Map
text(coordinates(HexPols2), labels=sapply(slot(HexPols2, "polygons"),
  function(i) slot(i, "ID")), cex=0.5)
```
\newline
14\. As an alternative to importing a polygon that we created in ArcMap, we can create a polygon in R using the coordinates of the boundary box of the area of interest. In our case here, the bounding box will be the mule deer locations.
```{r fig.height=4, fig.width=4, warning=FALSE, message=FALSE}
proj4string(deer.spdf)
bbox(deer.spdf@coords)
bb <- cbind(x=c(-108.83966,-108.83966,-108.9834,-108.9834, -108.83966), 
  y=c(37.8142, 37.86562,37.86562,37.8142,37.8142))
SP <- SpatialPolygons(list(Polygons(list(Polygon(bb)),"1")), proj4string=CRS(proj4string(MDclip)))
plot(SP)
proj4string(SP)
points(deer.spdf,col="red")
```
\newline
15\. Now make practical use of the new bounding box we created by clipping a larger raster dataset. A smaller raster dataset runs analyses faster, provides a zoomed in view of mule deer locations and vegetation, and is just easier to work with.
```{r message=FALSE, fig.height=4, fig.width=4}
#Load vegetation raster layer textfile clipped in ArcMap 
veg <-raster("extentnlcd2.txt")
plot(veg)
class(veg)
```
```{r eval=FALSE, warning=FALSE, message=FALSE}
#Clip using the raster imported with "raster" package
bbclip <- crop(veg, SP)
veg
```
```{r warning=FALSE, message=FALSE }
#WON'T WORK because projections are not the same, WHY?
#Let's check projections of layers we are working with now.
proj4string(MDclip)
proj4string(deer.spdf)
proj4string(SP)
proj4string(veg)
```
16\. We need to have all layers in same projection so project the deer.spdf to Albers and then clip vegetation layer with new polygon we created in the Albers projection.
```{r warning=FALSE, message=FALSE}
deer.albers <-spTransform(deer.spdf, CRS=Albers.crs)
class(deer.albers)
proj4string(deer.albers)

bbox(deer.albers)
bb1 <- cbind(x=c(-1115562,-1115562,-1127964,-1127964,-1115562),
  y=c(1718097, 1724867,1724867,1718097,1718097))
AlbersSP <- SpatialPolygons(list(Polygons(list(Polygon(bb1)),"1")), 
  proj4string=CRS(proj4string(deer.albers)))
plot(AlbersSP)
points(deer.albers)
#Check to see all our layers are now in Albers projection
proj4string(veg)
proj4string(deer.albers)
proj4string(AlbersSP)

#Clip using the raster imported with "raster" package 
bbclip <- crop(veg, AlbersSP)
plot(bbclip)
points(deer.albers, col="red")
plot(AlbersSP, lwd=5, add=T)
text(coordinates(AlbersSP), labels="Colorado Mule Deer")
```

